\documentclass{llncs}

\usepackage{makeidx}
\usepackage{amssymb}
\usepackage{listings}
\usepackage{indentfirst}
\usepackage{verbatim}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{url}
\usepackage{stmaryrd}
\usepackage{xspace}
\usepackage{comment}
\usepackage{wrapfig}
\usepackage{placeins}
\usepackage{tabularx}
\usepackage{ragged2e}
\usepackage{subcaption}
\captionsetup{compatibility=false}
%\usepackage{natbib}
%\usepackage [sorting = none] {biblatex}
%\addbibresource {main.bib}

\def\transarrow{\xrightarrow}
\newcommand{\setarrow}[1]{\def\transarrow{#1}}

\newcommand{\trule}[2]{\frac{#1}{#2}}
\newcommand{\crule}[3]{\frac{#1}{#2},\;{#3}}
\newcommand{\withenv}[2]{{#1}\vdash{#2}}
\newcommand{\trans}[3]{{#1}\transarrow{#2}{#3}}
\newcommand{\ctrans}[4]{{#1}\transarrow{#2}{#3},\;{#4}}
\newcommand{\llang}[1]{\mbox{\lstinline[mathescape]|#1|}}
\newcommand{\pair}[2]{\inbr{{#1}\mid{#2}}}
\newcommand{\inbr}[1]{\left<{#1}\right>}
\newcommand{\highlight}[1]{\color{red}{#1}}
\newcommand{\ruleno}[1]{\eqno[\scriptsize\textsc{#1}]}
\newcommand{\inmath}[1]{\mbox{$#1$}}
\newcommand{\lfp}[1]{fix_{#1}}
\newcommand{\gfp}[1]{Fix_{#1}}
\newcommand{\vsep}{\vspace{-2mm}}
\newcommand{\supp}[1]{\scriptsize{#1}}
\newcommand{\G}{\mathfrak G}
\newcommand{\sembr}[1]{\llbracket{#1}\rrbracket}
\newcommand{\cd}[1]{\texttt{#1}}
\newcommand{\miniKanren}{miniKanren\xspace}
\newcommand{\ocanren}{OCanren\xspace}
\newcommand{\free}[1]{\boxed{#1}}
\newcommand{\binds}{\;\mapsto\;}
\newcommand{\dbi}[1]{\mbox{\bf{#1}}}

\let\emptyset\varnothing
\setlength{\abovecaptionskip}{3pt plus 3pt minus 2pt}
\setlength{\belowcaptionskip}{-10pt plus 3pt minus 2pt}

\lstdefinelanguage{ocanren}{
keywords={fresh, let, in, match, with, when, class, type,
object, method, of, rec, repeat, until, while, not, do, done, as, val, inherit,
new, module, sig, deriving, datatype, struct, if, then, else, open, private, virtual, include, success, failure,
true, false},
sensitive=true,
commentstyle=\small\itshape\ttfamily,
keywordstyle=\ttfamily\underbar,
identifierstyle=\ttfamily,
basewidth={0.5em,0.5em},
columns=fixed,
fontadjust=true,
literate={fun}{{$\lambda$}}1 {->}{{$\to$}}3 {===}{{$\equiv$}}1 {=/=}{{$\not\equiv$}}1 {|>}{{$\triangleright$}}3 {|||}{{$\vee$}}2 {/\\}{{$\wedge$}}2 {^}{{$\uparrow$}}1,
morecomment=[s]{(*}{*)}
}

\lstset{
mathescape=true,
%basicstyle=\small,
identifierstyle=\ttfamily,
keywordstyle=\bfseries,
commentstyle=\scriptsize\rmfamily,
basewidth={0.5em,0.5em},
fontadjust=true,
language=ocanren
}

\usepackage{letltxmacro}
\newcommand*{\SavedLstInline}{}
\LetLtxMacro\SavedLstInline\lstinline
\DeclareRobustCommand*{\lstinline}{%
  \ifmmode
    \let\SavedBGroup\bgroup
    \def\bgroup{%
      \let\bgroup\SavedBGroup
      \hbox\bgroup
    }%
  \fi
  \SavedLstInline
}
%\addtolength{\parskip}{-2pt}
%\setlength{\parskip}{0pt}
\setlength{\belowcaptionskip}{-15pt}

\pagestyle{plain}
\begin{document}
\sloppy
\mainmatter

\title{Typed Relational Conversion\thanks{This work is supported by RFBR grant No 18-01-00380.}}

\author{
  Petr Lozov\inst{1} \and Andrei Vyatkin\inst{2} \and Dmitry Boulytchev\inst{3}
}

\institute{
St.Petersburg State University,\\
Universitetski pr., 28, 198504, St.Petersburg, Russia,\\
\email{lozov.peter@gmail.com}
\and
St.Petersburg State University,\\
Universitetski pr., 28, 198504, St.Petersburg, Russia,\\
\email{dewshick@gmail.com}
\and
St.Petersburg State University,\\
Universitetski pr., 28, 198504, St.Petersburg, Russia,\\
JetBrains Research\\
Universitetskaya emb., 7-9-11, bldg. 5A, 199034, St.Petersburg, Russia,\\
\email{dboulytchev@math.spbu.ru}}

\maketitle

\begin{abstract}
We address the problem of transforming typed functional programs into relational form. 
In this form, a program can be run in various ``directions'' with some arguments left free, 
making it possible to acquire different behaviors from a single specification. We specify the 
syntax, typing rules and semantics for the source language as well as its relational extension, 
describe the conversion and prove its correctness both in terms of typing and dynamic semantics. 
We also discuss the limitations of our approach, present the implementation of the conversion for 
the subset of OCaml and evaluate it on a number of realistic examples.
\end{abstract}

\input{introduction.tex}
\input{ocanren.tex}
\input{language.tex}
\input{conversion.tex}
\input{evaluation.tex}
\input{conclusion.tex}

%\begin{comment}
\begin{thebibliography}{10}

\bibitem{CKanren}
C.~E. Alvis, J.~J. Willcock, K.~M. Carter, W.~E. Byrd, and D.~P. Friedman.
\newblock {cKanren}: {miniKanren} with Constraints.
\newblock In {\em Proceedings of the 2011 Annual Workshop on Scheme and
  Functional Programming}, Oct. 2011.

\bibitem{Lambda}
H.~P. Barendregt.
\newblock Lambda Calculi with Types.
\newblock In {\em Handbook of Logic in Computer Science (vol. 2)}, 
pages 117--309. Oxford University Press, Inc., New York, NY, USA, 1992.

\bibitem{WillOnHM}
W.~E. Byrd.
\newblock Private communications.

\bibitem{WillThesis}
W.~E. Byrd.
\newblock Relational Programming in miniKanren: Techniques, Applications,
  and Implementations.
\newblock PhD thesis, Indiana University, September 2009.

\bibitem{unified}
W.~E. Byrd, M.~Ballantyne, G.~Rosenblatt, and M.~Might.
\newblock A Unified Approach to Solving Seven Programming Problems (functional
  pearl).
\newblock {\em Proc. ACM Program. Lang.}, 1(ICFP):8:1--8:26, Aug. 2017.

\bibitem{alphaKanren}
W.~E. Byrd and D.~P. Friedman.
\newblock {$\alpha$Kanren}: A Fresh Name in Nominal Logic Programming.
\newblock In {\em Proceedings of the 2007 Annual Workshop on Scheme and
  Functional Programming}, pages 79--90, 2007.

\bibitem{Untagged}
W.~E. Byrd, E.~Holk, and D.~P. Friedman.
\newblock miniKanren, Live and Untagged: Quine Generation via Relational
  Interpreters (programming pearl).
\newblock In {\em Proceedings of the 2012 Annual Workshop on Scheme and
  Functional Programming}, Scheme '12, pages 8--29, New York, NY, USA, 2012.
  ACM.

\bibitem{cardelli}
L.~Cardelli and P.~Wegner.
\newblock On Understanding Types, Data Abstraction, and Polymorphism.
\newblock {\em ACM Comput. Surv.}, 17(4):471--523, Dec. 1985.

\bibitem{TRS}
D.~P. Friedman, W.~E. Byrd, and O.~Kiselyov.
\newblock The Reasoned Schemer.
\newblock The MIT Press, 2005.

\bibitem{MicroKanren}
J.~Hemann and D.~P. Friedman.
\newblock $\mu$Kanren: A Minimal Functional Core for Relational Programming.
\newblock In {\em Proceedings of the 2013 Annual Workshop on Scheme and
  Functional Programming}, 2013.

\bibitem{SmallEmbedding}
J.~Hemann, D.~P. Friedman, W.~E. Byrd, and M.~Might.
\newblock A Small Embedding of Logic Programming with a Simple Complete Search.
\newblock {\em SIGPLAN Not.}, 52(2):96--107, Nov. 2016.

\bibitem{ocanren}
D.~Kosarev and D.~Boulytchev.
\newblock Typed Embedding of a Relational Language in OCaml.
\newblock {\em ACM SIGPLAN Workshop on ML}, 2016.

\bibitem{UnificationRevisited}
J.-L. Lassez, M.~J. Maher, and K.~Marriott.
\newblock Unification Revisited.
\newblock In {\em Foundations of Deductive Databases and Logic Programming},
pages 587--625. Morgan Kaufmann Publishers Inc., San Francisco, CA, USA, 1988.

\bibitem{Types}
B.~C. Pierce.
\newblock Types and Programming Languages.
\newblock The MIT Press, 1st edition, 2002.

\bibitem{Unification}
F.~Baader and W.~Snyder. 
\newblock{Unification Theory.}
\newblock In {\em Handbook of Automated Reasoning},
Elsevier Science Publishers B. V., Amsterdam, The Netherlands, The Netherlands, 2001.

\bibitem{Felleisen}
A.~Wright and M.~Felleisen.
\newblock A Syntactic Approach to Type Soundness.
\newblock {\em Inf. Comput.}, 115(1):38--94, Nov. 1994.

\end{thebibliography}
%\end{comment}

%\renewcommand{\clearpage}{} 
%\bibliographystyle{abbrv}
%\bibliography{main}

%\clearpage
%\appendix
%\input{appendix.tex}

\end{document}

