\section{Streams, States, and Goals}
\label{sec:goals}

This section describes a top-level framework for our implementation. Despite it contains
nothing more, than a reiteration of the original implementation~\cite{MicroKanren, CKanren}
in OCaml, we still need some notions to be properly established.

\miniKanren is organized as a set of combinators, designated to describe a search for
the solution of a certain \emph{goal}. The search itself is implemented using a
backtracking lazy stream monad:

\begin{lstlisting}
   type $\alpha$ stream

   val mplus : $\alpha$ stream -> $\alpha$ stream -> $\alpha$ stream
   val bind  : $\alpha$ stream -> ($\alpha$ -> $\beta$ stream) -> $\beta$ stream
\end{lstlisting}

Monadic primitives describe the shape of the search, and their implementations may
vary in concrete \miniKanren versions.

An essential component of the implementation is a bundle of the following types:

\begin{lstlisting}
   type env         = $\dots$
   type subst       = $\dots$
   type constraints = $\dots$

   type state = env * subst * constraints
\end{lstlisting}

Type \lstinline{state} describes a point in a lazily constructed search tree: type \lstinline{env} corresponds
to an \emph{environment}, which contains some supplementary information (in particular, environment is needed to
correctly allocate fresh variables), type \lstinline{subst} describes a substitution, which keeps current bindings
for some logical variables, and, finally, type \lstinline{constraints} represents disequality constraints,
which have to be respected. In a simplest case \lstinline{env} is just a counter for the number of the next free
variable, \lstinline{subst} is a map-like structure and \lstinline{constraints} is a list of substitutions. In our
case environment contains some extra information to make it possible to identify variables in a constant time.

The next cornerstone element is the \emph{goal} type, which is considered as a transformer of a state into
lazy stream of states:

\begin{lstlisting}
   type goal = state -> state stream
\end{lstlisting}

In terms of search, a goal nondeterministically performs one step of the search: for a given
node in a search tree it produces its immediate descendants. On the user level type \lstinline{goal}
is abstract, and states are completely hidden.

Next to last, there is a number of predefined combinators:

\begin{lstlisting}
   val (&&&)      : goal -> goal -> goal
   val (|||)      : goal -> goal -> goal
   val call_fresh : ($v$ -> goal) -> goal
   ....
\end{lstlisting}

Conjunction ``\lstinline{&&&}'' combines the results of its argument goals using \lstinline{bind},
disjunction ``\lstinline{|||}'' concatenates the results using \lstinline{mplus}, abstraction
primitive \lstinline{call_fresh} takes an abstracted goal and applies it to a freshly created
variable. Type $v$ in the latest case designates the type for a fresh variable, which we leave
abstract for now. These combinators serve as bricks for implementation of conventional
\miniKanren constructions and syntax extensions (\lstinline{conde}, \lstinline{fresh} etc.)

Finally, there are two primitive goal constructors:

\begin{lstlisting}
   val (===) : $t$ -> $t$ -> goal
   val (=/=) : $t$ -> $t$ -> goal
\end{lstlisting}

The first one is unification, while the other~--- disequality constraint. Here we, again, left
the type of terms $t$ abstract; it will be instantiated later.

In the implementation of \miniKanren both these goals are implemented using unification~\cite{CKanren}; this
is true for us as well. However, due to a drastic difference in host languages the implementation of
efficient polymorphic unification itself leads to a number of tricks with typing and data representation,
non-existing in the original version.

Described combinators allow to \emph{construct} goals. To \emph{run} a goal, the following
primitive is used:

\begin{lstlisting}
   val run : goal -> state stream
\end{lstlisting}

This primitive creates an initial state and applies a goal to it. The states in the return stream describe
various solutions for the goal. As the stream is constructed lazily, taking elements one by one makes
the search progress.

To discover concrete answers, the state has to be queried for its contents. As a rule, a few variables
are \emph{refined} in a state, i.e. their bindings in corresponding substitution are retrieved. For free
variables in addition their disequality constraints have to be \emph{reified} (e.g. represented as a list of
``forbidden'' terms). As forbidden terms can in turn contain free variables, the constraint reification
process is recursive.

In our case refinement and reification are subtle parts, since, as we will see shortly, they can not be
implemented in a type-safe part of the language.
