\section{Related Work}

The study of formal semantics for logic programming languages, particularly \textsc{Prolog}, is a well-established research domain. Early
works~\cite{JonesMycroftSemantics,DebrayMishraSemantics} addressed the computational aspects of both pure \textsc{Prolog} and its extension
with the cut construct. Recently, the application of certified/mechanized approaches came into focus as well. In particular,
in one work~\cite{CertifiedPrologEquivalences} the equivalence of a few differently defined semantics
for pure \textsc{Prolog} is proven, and in another work~\cite{CeritfiedDenotationalCut} a denotational semantics for \textsc{Prolog} with cut is presented; both
works provide \textsc{Coq}-mechanized proofs. It is interesting that the former one also advocates the use of higher-order
abstract syntax. We are not aware of any prior work on certified semantics for \textsc{Prolog} which contributed a correct-by-construction
interpreter. Our certified description of SLD resolution with cut can be considered as a certified semantics for \textsc{Prolog} modulo
occurs check in unification (which \textsc{Prolog} does not have by default).

The implementation of first-order unification in dependently typed languages constitutes a well-known challenge with a number of
known solutions. The major difficulty comes from the non-structural recursivity of conventional unification algorithms, which
requires to provide a witness for convergence. The standard approach is to define a generally-recursive function and a well-founded order
for its arguments. This route is taken in a number of works~\cite{MGUinLCF,MGUinMLTT,IdempMGUinCoq,TextbookMGUinCoq}, where the descriptions of
unification algorithms are given in \textsc{Coq}, \textsc{LCF} and \textsc{Alf}. The well-founded used there is
lexicographically ordered tuples, containing the information about the number of different free variables and the sizes of
the arguments. We implement a similar approach, but we separate the test for the non-matching case into a dedicated
function. Thus, we make a recursive call only when the current substitution extension is guaranteed, which allows us to use the
number of different free variables as the well-founded order. An alternative approach suggested by McBride~\cite{StructuralMGU} gives a structurally recursive definition of
the unification algorithm; this is achieved by indexing the arguments with the numbers of their free variables.

The use of higher-order abstract syntax (HOAS) for dealing with language constructs in \textsc{Coq} was addressed in early work~\cite{HOASinCoq},
where it was employed to describe the lambda calculus. The inconsistency phenomenon of HOAS representation, mentioned in Section~\ref{specification}, is called
there ``exotic terms'' there and is handled using a dedicated inductive predicate ``\lstinline|Valid_v|''. The predicate has a non-trivial implementation based
on subtle observations on the behavior of bindings. Our case, however, is much simpler: there is not much variety in ``exotic terms'' (for example, we do not have
reductions in terms), and our consistency predicate can be considered as a limited version of ``\lstinline|Valid_v|'' for a smaller language.

The study of formal semantics for \textsc{miniKanren} is not a completely novel venture. Previously, a nondeterministic
small-step semantics was described~\cite{RelConversion}, as well as a big-step semantics for a finite number of answers~\cite{DivTest};
neither uses proof mechanization and in both works the interleaving is not addressed. 

The work of Kumar~\cite{MechanisingMiniKanren} can be considered as our direct predecessor. It also introduces both denotational and
operational semantics and presents a \textsc{HOL}-certified proof for the soundness of the latter w.r.t. the former. The denotational
semantics resembles ours but considers only queries with a single free variable (we do not see this restriction as important).
On the other hand, the operational semantics is non-deterministic, which makes it
impossible to express interleaving and extract the interpreter in a direct way. In addition, a specific form of ``executable semantics''
is introduced, but its connection to the other two is not established. Finally, no completeness result is presented.
We consider our completeness proof as an essential improvement. 

The most important property of interleaving search~--- completeness~--- was postulated in the introductory paper~\cite{Search}, and is delivered by
all major implementations. Hemann et al.~\cite{SmallEmbedding} give a proof of completeness for a specific implementation of \textsc{miniKanren};
however, the completeness is understood there as
preservation of all answers during the interleaving of answer streams, i.e. in a more narrow sense than in our work since no relation
to denotational semantics is established.
