\begin{figure}[t]
\[
\begin{array}{rcl}
  x\,[t/x] &=& t \\
  y\,[t/x] &=& y,\;\; y\ne x\\
  C_i^{k_i}\,(t_1,\dots,t_{k_i})\,[t/x]&=&C_i^{k_i}\,(t_1\,[t/x],\dots,t_{k_i}\,[t/x])\\
  (t_1 \equiv t_2)\,[t/x]&=&t_1\,[t/x] \equiv t_2\,[t/x]\\
  (g_1 \wedge g_2)\,[t/x]&=&g_1\,[t/x] \wedge g_2\,[t/x]\\
  (g_1 \vee g_2)\,[t/x]&=&g_1\,[t/x] \vee g_2\,[t/x]\\
  (\mbox{\lstinline|fresh|}\;x\,.\,g)\,[t/x]&=&\mbox{\lstinline|fresh|}\;x\,.\,g\\
  (\mbox{\lstinline|fresh|}\;y\,.\,g)\,[t/x]&=&\mbox{\lstinline|fresh|}\;y\,.\,(g\,[t/x]),\;\;y\ne x\\
  (R_i^{k_i}\,(t_1,\dots,t_{k_i})\,[t/x]&=&R_i^{k_i}\,(t_1\,[t/x],\dots,t_{k_i}\,[t/x])
\end{array}
\]
  \caption{Substitutions for terms and goals}
  \label{substitution}
\end{figure}

\section{Denotational Semantics}
\label{denotational}

In this section we present a denotational semantics for the language we defined above. We use a simple set-theoretic
approach analogous to the least Herbrand model for definite logic programs~\cite{LHM}.
Strictly speaking, instead of developing it from scratch we could have just described the conversion of specifications
into definite logic form and took their least Herbrand model. However, in that case, we would still need to define
the least Herbrand model semantics for definite logic programs in a certified way. In addition, while for
this concrete language the conversion to definite logic form is trivial, it may become less trivial for
its extensions (with, for example, nominal constructs~\cite{AlphaKanren}) which we plan to do in future.

We also must make the following observations. First, building inductive denotational semantics in a conventional way amounts to
constructing a complete lattice and a monotone function and taking its least fixed point~\cite{TarskiKnaster}.
As we deal with a first-order language with only monotonic constructs (conjunction/disjunction) these steps
are trivial. Moreover, we express the semantics in \textsc{Coq}, where all well-formed inductive definitions already
have proper semantics, which removes the necessity to justify the validity of the steps we perform. Second, 
the least Herbrand model is traditionally defined as the least fixed point of a transition function (defined by a logic program)
which maps sets of ground atoms to sets of ground atoms. We are, however, interested in \emph{relational} semantics which
should map a program into $n$-ary relation over ground terms, where $n$ is the number of free variables in the topmost
goal. Thus, we deviate from the traditional route and describe the denotational semantics in a more specific way.

To motivate further development, we first consider the following example. Let us have the following goal:

\begin{lstlisting}
   x === Cons (y, z)
\end{lstlisting}

There are three free variables, and solving the goal delivers us the following single answer:

\begin{lstlisting}
   $\alpha\mapsto\;$ Cons ($\beta$, $\gamma$)
\end{lstlisting}

where semantic variables $\alpha$, $\beta$ and $\gamma$ correspond to the syntactic ones ``\lstinline|x|'', ``\lstinline|y|'', ``\lstinline|z|''. The
goal does not put any constraints on ``\lstinline|y|'' and ``\lstinline|z|'', so there are no bindings for ``$\beta$'' and ``$\gamma$'' in the answer.
This answer can be seen as the following ternary relation over the set of all ground terms:

\[
\{(\mbox{\lstinline|Cons ($\beta$, $\,\gamma$)|}, \beta, \gamma) \mid \beta\in\mathcal{D},\,\gamma\in\mathcal{D}\}\subset\mathcal{D}^3
\]

The order of ``dimensions'' is important, since each dimension corresponds to a certain free variable. Our main idea is to represent this relation as a set of total
functions 

\[
\mathfrak{f}:\mathcal{A}\mapsto\mathcal{D}
\]

from semantic variables to ground terms. We call these functions \emph{representing functions}. Thus, we may reformulate the same relation as

\[
\{(\mathfrak{f}\,(\alpha),\mathfrak{f}\,(\beta),\mathfrak{f}\,(\gamma))\mid\mathfrak{f}\in\sembr{\mbox{\lstinline|$\alpha$ === Cons ($\beta$, $\,\gamma$)|}}\}
\]

where we use conventional semantic brackets ``$\sembr{\bullet}$'' to denote the semantics. For the top-level goal, we need to substitute its free syntactic
variables with distinct semantic ones, calculate the semantics, and build the explicit representation for the relation as shown above. The relation, obviously,
does not depend on the concrete choice of semantic variables but depends on the order in which the values of representing functions are tupled. This order can be
conventionalized, which gives us a completely deterministic semantics.

Now we implement this idea. First, for a representing function

\[
\mathfrak{f} : \mathcal{A}\to\mathcal{D}
\]

we introduce its homomorphic extension 

\[
  \overline{\mathfrak{f}}:\mathcal{T_A}\to\mathcal{D}
\]

which maps terms to terms:

\[
\begin{array}{rcl}
  \overline{\mathfrak f}\,(\alpha) & = & \mathfrak f\,(\alpha)\\
  \overline{\mathfrak f}\,(C_i^{k_i}\,(t_1,\dots.t_{k_i})) & = & C_i^{k_i}\,(\overline{\mathfrak f}\,(t_1),\dots \overline{\mathfrak f}\,(t_{k_i}))
\end{array}
\]

{ \color{red}

Let us have two terms $t_1, t_2\in\mathcal{T_A}$. If there is a unifier for $t_1$ and $t_2$ then, clearly, there is a substitution $\theta$ which
turns both $t_1$ and $t_2$ into the same \emph{ground} term (we do not require $\theta$ to be the most general). Thus, $\theta$ maps
(some) variables into ground terms, and its application to $t_{1(2)}$ is exactly $\overline{\theta}(t_{1(2)})$. This reasoning can be
performed in the opposite direction: a unification $t_1\equiv t_2$ defines the set of all representing functions $\mathfrak{f}$ for which
$\overline{\mathfrak{f}}(t_1)=\overline{\mathfrak{f}}(t_2)$.

}

We remind the conventional notions of pointwise modification of a function

\[
f\,[x\gets v]\,(z)=\left\{
\begin{array}{rcl}
  f\,(z) &,& z \ne x \\
  v      &,& z = x
\end{array}
\right.
\]

and substitution of a free variable with a term in terms and goals (see Figure~\ref{substitution}).

For a representing function $\mathfrak{f}:\mathcal{A}\to\mathcal{D}$ and a semantic variable $\alpha$ we define
the following \emph{generalization} operation:

\[
\mathfrak{f}\uparrow\alpha = \{ \mathfrak{f}\,[\alpha\gets d] \mid d\in\mathcal D\}
\]

Informally, this operation generalizes a representing function into a set of representing functions in such a way that the
values of these functions for a given variable cover the whole $\mathcal{D}$. We extend the generalization operation for sets of
representing functions $\mathfrak{F}\subseteq\mathcal{A}\to\mathcal{D}$:

\[
  \mathfrak{F}\uparrow\alpha = \bigcup_{\mathfrak{f}\in\mathfrak{F}}(\mathfrak{f}\uparrow\alpha)
\]

Now we are ready to specify the semantics for goals (see Figure~\ref{denotational_semantics_of_goals}).
 We've already given the motivation for
  the semantics of unification: the condition $\overline{\mathfrak{f}}(t_1)=\overline{\mathfrak{f}}(t_2)$ gives us the set of all (otherwise
  unrestricted) representing functions which ``equate'' terms $t_1$ and $t_2$.
  Set union and intersection provide a conventional interpretation
for disjunction and conjunction of goals. Now for the case of ``\lstinline|fresh $x$ . $g$|''. First, we take an arbitrary semantic variable $\alpha$,
not free in $g$, and substitute $x$ with $\alpha$. Then we calculate the semantics of $g\,[\alpha/x]$. The interesting part is the next step:
as $x$ can not be free in ``\lstinline|fresh $x$ . $g$|'', we need to generalize the result over $\alpha$ since in our model the semantics of a
goal specifies a relation over its free variables. We introduce some nondeterminism by choosing arbitrary $\alpha$, but we can prove that with different
choices of free variable the semantics of a goal does not change:

\begin{lemma}
\label{ChangeFreeVarLemma}
For any goal 

\[
\mbox{\lstinline|fresh|}\;x\,.\,g
\]

for any two variables $\alpha$ and $\beta$ which are not free in this goal,
if $\mathfrak{f} \in \sembr{g\,[\alpha/x]}$, then for any representing function $\mathfrak{f}'$, such that

\begin{enumerate}
\item $\mathfrak{f}'(\beta) = \mathfrak{f}(\alpha)$,
\item $\forall \gamma: \gamma \neq \alpha \land \gamma \neq \beta,\; \mathfrak{f}'(\gamma) = \mathfrak{f}(\gamma)$,
\end{enumerate}

it is true that

\[
\mathfrak{f}' \in \sembr{g\,[\beta/x]}
\]
\end{lemma}
%\begin{proof}
  The proof turned out to be the most cumbersome among all others. \orangenote{\textbf{Более подробное доказательство в аппендикс, а здесь вставить:}
, in case of nested ‘fresh’, details may be found in the appendix.} {\color{orange} In the case of nested ``\lstinline|fresh|'' we have to construct two
  intermediate representing functions with different values on three points (one of which is non-free variable which is not equal to
  any of those we have, so we also have to prove that there is always exists ``other fresh variable'') and then examine the four cases
  with different relations (equal or not equal) between substituted variables. We also have to use the goal consistency condition here. }
%\end{proof}

\begin{comment} We introduce some nondeterminism by choosing arbitrary $\alpha$, but we have proven by
structural induction that with different choices of free variable the semantics of a goal does not change (and this proof turned out to
be the most cumbersome among all others). 
\end{comment}

\orangenote{\textbf{Поместить перед семантикой fresh}}
{\color{orange}

The final case is a relational invocation, in which we unfold the definition of the corresponding
relational symbol and substitute its formal parameters with actual ones.

}

\begin{figure}[t]
  \[
  \begin{array}{cclr}
    \sembr{t_1\equiv t_2}&=&\{\mathfrak f : \mathcal{A}\to\mathcal{D}\mid \overline{\mathfrak{f}}\,(t_1)=\overline{\mathfrak{f}}\,(t_2)\}& \ruleno{Unify$_D$}\\
    \sembr{g_1\wedge g_2}&=&\sembr{g_1}\cap\sembr{g_1}&\ruleno{Conj$_D$}\\
    \sembr{g_1\vee g_2}&=&\sembr{g_1}\cup\sembr{g_1}&\ruleno{Disj$_D$}\\
    \sembr{\mbox{\lstinline|fresh|}\,x\,.\,g}&=&(\sembr{g\,[\alpha/x]})\uparrow\alpha,\;\alpha\not\in FV(g)& \ruleno{Fresh$_D$}\\
    \sembr{R\,(t_1,\dots,t_k)}&=&\sembr{g\,[t_1/x_1,\dots,t_k/x_k]},\;\mbox{where}\;R=\lambda\,x_1\dots x_k\,.\,g & \ruleno{Invoke$_D$}
  \end{array}
  \]
  \caption{Denotational semantics of goals}
  \label{denotational_semantics_of_goals}
\end{figure}

Here is an example of denotational semantics of a goal:

%\renewcommand{\overset}[2]{#2}
\[
\begin{array}{lc}
  \sembr{\mbox{\lstinline|fresh y . ($\alpha$ === y) $\,\wedge\,$ (y === Zero)|}}&\overset{\mbox{(by \textsc{Fresh$_D$})}}{=}\\
  (\sembr{\mbox{\lstinline|($\alpha$ === $\beta$) $\,\wedge\,$ ($\beta$ === Zero)|}})\uparrow\beta&\overset{\mbox{(by \textsc{Conj$_D$})}}{=}\\
  (\sembr{\mbox{\lstinline|$\alpha$ === $\beta$|}} \,\cap\, \sembr{\mbox{\lstinline|$\beta$ === Zero)|}})\uparrow\beta&\overset{\mbox{(by \textsc{Unify$_D$})}}{=}\\
  (\{\mathfrak{f}\mid \overline{\mathfrak{f}}\,(\alpha)=\overline{\mathfrak{f}}\,(\beta)\} \,\cap\, \{\mathfrak{f}\mid \overline{\mathfrak{f}}\,(\beta)=\overline{\mathfrak{f}}\,(\mbox{\lstinline|Zero|})\})\uparrow\beta&\overset{\mbox{(by the definition of ``$\overline{\mathfrak{f}}$'')}}{=}\\
  (\{\mathfrak{f}\mid \mathfrak{f}\,(\alpha)=\mathfrak{f}\,(\beta)\} \,\cap\, \{\mathfrak{f}\mid \mathfrak{f}\,(\beta)=\mbox{\lstinline|Zero|}\})\uparrow\beta&\overset{\mbox{(by the definition of ``$\cap$'')}}{=}\\
  (\{\mathfrak{f}\mid \mathfrak{f}\,(\alpha)=\mathfrak{f}\,(\beta)=\mbox{\lstinline|Zero|}\})\uparrow\beta&\overset{\mbox{(by the definition of ``$\uparrow$'')}}{=}\\
  \{\mathfrak{f}\mid \mathfrak{f}\,(\alpha)=\mbox{\lstinline|Zero|}, \mathfrak{f}\,(\beta)=d, d\in\mathcal{D}\}&\overset{\mbox{(by the totality of representing functions)}}{=}\\[1mm]
  \{\mathfrak{f}\mid \mathfrak{f}\,(\alpha)=\mbox{\lstinline|Zero|}\}&
\end{array}
\]

In the end, we've got the set of representing functions, each of which restricts only the value of the free variable $\alpha$.

To formalize denotational semantics in \textsc{Coq} we can define representing functions simply as \textsc{Coq} functions:

\begin{lstlisting}[language=Coq]
   Definition repr_fun : Set := name -> ground_term.
\end{lstlisting}

We define the semantics via the inductive proposition ``\lstinline|in_denotational_sem_goal|'' (with notation ``\lstinline[mathescape=true]{[| $\bullet$ , $\bullet$ |]}'')
such that

\[
\forall g,\mathfrak{f}\;:\;\mbox{\lstinline|in_denotational_sem_goal|}\;g\;\mathfrak{f}\Longleftrightarrow\mathfrak{f}\in\sembr{g}
\]

\begin{comment}
The head of the definition is as follows

\begin{lstlisting}[language=Coq,morekeywords={where,at,level}]
  Reserved Notation "[| g , f |]" (at level 0).
  Inductive in_denotational_sem_goal :
    goal -> repr_fun -> Prop :=
    ...
  where "[| g , f |]" := (in_denotational_sem_goal g f).
\end{lstlisting}

and the body just goes through the cases shown in Fig.~\ref{denotational_semantics_of_goals}.
\end{comment}
  
We formulate the following important \emph{completeness condition} for the semantics of a goal $g$:

\begin{lemma}[Completeness condition]
\label{CompletenessConditionLemma}
For any goal $g$ and two representing functions ${\mathfrak f}$ and ${\mathfrak f'}$, such that $\left.{\mathfrak f}\right|_{FV(g)} = \left.{\mathfrak f'}\right|_{FV(g)}$
\[ {\mathfrak f} \in \sembr{g} \Leftrightarrow {\mathfrak f'} \in \sembr{g} \]
\end{lemma}
\begin{proof}
  The proof is by induction on evidence. The only non-trivial case is when the goal is a ``\lstinline|fresh|'' construct, there we need to change the value of
  the given function in one point to apply the inductive hypothesis.
\end{proof}

\begin{comment}
\[
\forall {\mathfrak f}, {\mathfrak f'}:  \left.{\mathfrak f}\right|_{FV(g)} = \left.{\mathfrak f'}\right|_{FV(g)}, \quad {\mathfrak f} \in \sembr{g} \Leftrightarrow {\mathfrak f'} \in \sembr{g}
\]
\end{comment}

In other words, representing functions for a goal $g$ restrict only the values of free variables of $g$ and do not introduce any ``hidden'' correlations.
This condition guarantees that our semantics is complete in the sense that it does not introduce artificial restrictions for the relation it defines.

\begin{comment}
We prove that the semantics of goals always satisfies this condition:

The following lemma proves that the semantics of goals always satisfies this condition:

\begin{lstlisting}[language=Coq]
  Lemma completeness_condition (f f' : repr_fun) (g : goal),
    (forall x, is_fv_of_goal x g -> gt_eq (f x) (f' x)) ->
    [| g , f |] ->
    [| g , f' |].
\end{lstlisting}

Here ``\lstinline[language=Coq]{gt_eq}'' stands for the equality of ground terms.
\end{comment}
